---
title: "Bird and Wildlife Diversity Correlation Analysis Using 30-Minute Independent Events"
author: "Jacqueline Wu"
date: "2026-04-16"
format:
  html:
    toc: true
    embed-resources: true
---

```{r setup, message=FALSE, warning=FALSE}
# Load packages
library(tidyverse)
library(lubridate)
library(vegan)
library(knitr)
```

#### 1. Read data
```{r}
birds <- read_csv("E:/TNC/detections_kept_gte_threshold.csv")
wildlife <- read_csv("E:/TNC/wildlife insights/images_2009899.csv")
```
#### 2. Clean and standardize data
```{r}
birds_clean <- birds %>%
  mutate(
    site = as.character(site),
    plot = as.numeric(plot),
    species = str_to_lower(detection_common_name),
    date = as.Date(date),
    hour = as.numeric(hour),
    start_s = as.numeric(start_s),
    datetime = as.POSIXct(date) + hours(hour) + seconds(start_s)
  ) %>%
  filter(
    !is.na(site),
    !is.na(plot),
    !is.na(species),
    !is.na(datetime)
  )

wildlife_clean <- wildlife %>%
  mutate(
    deployment_id = as.character(deployment_id),
    site = str_extract(deployment_id, "^[^_]+"),
    plot = str_extract(deployment_id, "(?<=_)[0-9]+") %>% as.numeric(),
    species = str_to_lower(common_name),
    timestamp = parse_date_time(timestamp, orders = c("Y/m/d H:M", "Y/m/d H:M:S")),
    number_of_objects = as.numeric(number_of_objects)
  ) %>%
  filter(
    !is.na(site),
    !is.na(plot),
    !is.na(species),
    !is.na(timestamp)
  )
```
#### 2b. compare based on calendar days
```{r}
birds_clean <- birds %>%
  mutate(
    site = as.character(site),
    plot = as.numeric(plot),
    species = str_to_lower(detection_common_name),
    date = as.Date(date),
    hour = as.numeric(hour),
    start_s = as.numeric(start_s),
    datetime = as.POSIXct(date) + hours(hour) + seconds(start_s),
    sample_day = as.Date(datetime)
  ) %>%
  filter(
    !is.na(site),
    !is.na(plot),
    !is.na(species),
    !is.na(datetime),
    !is.na(sample_day)
  )

wildlife_clean <- wildlife %>%
  mutate(
    deployment_id = as.character(deployment_id),
    site = str_extract(deployment_id, "^[^_]+"),
    plot = str_extract(deployment_id, "(?<=_)[0-9]+") %>% as.numeric(),
    species = str_to_lower(common_name),
    timestamp = parse_date_time(str_trim(as.character(timestamp)),
                                orders = c("Y/m/d H:M", "Y/m/d H:M:S")),
    number_of_objects = as.numeric(number_of_objects),
    sample_day = as.Date(timestamp)
  ) %>%
  filter(
    !is.na(site),
    !is.na(plot),
    !is.na(species),
    !is.na(timestamp),
    !is.na(sample_day)
  )
```
#### 3.  Keep only matched plots wildlife is the reference
```{r}
bird_plots <- birds_clean %>%
  distinct(site, plot)

wild_plots <- wildlife_clean %>%
  distinct(site, plot)

matched_plots <- inner_join(wild_plots, bird_plots, by = c("site", "plot")) %>%
  arrange(site, plot)

# Keep only matched plots in both datasets
birds_matched <- birds_clean %>%
  semi_join(matched_plots, by = c("site", "plot"))

wildlife_matched <- wildlife_clean %>%
  semi_join(matched_plots, by = c("site", "plot"))
```
#### 3b. find common days
```{r}
# Shared plot-day combinations
bird_plot_days <- birds_clean %>%
  semi_join(matched_plots, by = c("site", "plot")) %>%
  distinct(site, plot, sample_day)

wild_plot_days <- wildlife_clean %>%
  semi_join(matched_plots, by = c("site", "plot")) %>%
  distinct(site, plot, sample_day)

matched_plot_days <- inner_join(
  bird_plot_days,
  wild_plot_days,
  by = c("site", "plot", "sample_day")
) %>%
  arrange(site, plot, sample_day)

# Keep only shared plot-days
birds_daymatched <- birds_clean %>%
  semi_join(matched_plot_days, by = c("site", "plot", "sample_day"))

wildlife_daymatched <- wildlife_clean %>%
  semi_join(matched_plot_days, by = c("site", "plot", "sample_day"))
```
#### 4. Create 30-minute independent windows for wildlife
```{r}
wildlife_events <- wildlife_matched %>%
  arrange(site, plot, species, timestamp) %>%
  group_by(site, plot, species) %>%
  mutate(
    time_diff_min = as.numeric(difftime(timestamp, lag(timestamp), units = "mins")),
    new_event = if_else(is.na(time_diff_min) | time_diff_min > 30, 1, 0),
    event_id = cumsum(new_event)
  ) %>%
  ungroup()

# Keep one row per independent event
wildlife_independent <- wildlife_events %>%
  distinct(site, plot, species, event_id)
```
#### 5. Create 30-minute windows for birds
```{r}
# For birds, we use datetime and collapse detections into 30-minute bins.
# This is not exactly the same as camera-trap independent events,
# but it makes the temporal unit more comparable.

birds_30min <- birds_matched %>%
  mutate(
    window_30min = floor_date(datetime, unit = "30 minutes")
  ) %>%
  distinct(site, plot, species, window_30min)
```
#### 4b+5b. 30-minute windows for wildlife and birds
```{r}
wildlife_events_day <- wildlife_daymatched %>%
  arrange(site, plot, sample_day, species, timestamp) %>%
  group_by(site, plot, sample_day, species) %>%
  mutate(
    time_diff_min = as.numeric(difftime(timestamp, lag(timestamp), units = "mins")),
    new_event = if_else(is.na(time_diff_min) | time_diff_min > 30, 1, 0),
    event_id = cumsum(new_event)
  ) %>%
  ungroup()

wildlife_independent_day <- wildlife_events_day %>%
  distinct(site, plot, sample_day, species, event_id)

birds_30min_day <- birds_daymatched %>%
  mutate(
    window_30min = floor_date(datetime, unit = "30 minutes")
  ) %>%
  distinct(site, plot, sample_day, species, window_30min)
```
#### 6. Build community matrices by plot
```{r}
# ----- Bird matrix -----
bird_comm <- birds_30min %>%
  count(site, plot, species, name = "abundance") %>%
  unite("plot_id", site, plot, remove = FALSE) %>%
  select(plot_id, species, abundance) %>%
  pivot_wider(
    names_from = species,
    values_from = abundance,
    values_fill = 0
  )

bird_mat <- bird_comm %>%
  column_to_rownames("plot_id")

# ----- Wildlife matrix -----
wild_comm <- wildlife_independent %>%
  count(site, plot, species, name = "abundance") %>%
  unite("plot_id", site, plot, remove = FALSE) %>%
  select(plot_id, species, abundance) %>%
  pivot_wider(
    names_from = species,
    values_from = abundance,
    values_fill = 0
  )

wild_mat <- wild_comm %>%
  column_to_rownames("plot_id")

# Make sure plot order is aligned
common_plot_ids <- intersect(rownames(bird_mat), rownames(wild_mat))

bird_mat <- bird_mat[common_plot_ids, , drop = FALSE]
wild_mat <- wild_mat[common_plot_ids, , drop = FALSE]

cat("Number of common plot IDs used in diversity comparison:", length(common_plot_ids), "\n")
```
#### 7. Calculate plot-level diversity
```{r}
bird_div <- tibble(
  plot_id = rownames(bird_mat),
  bird_richness = specnumber(bird_mat),
  bird_shannon = diversity(bird_mat, index = "shannon"),
  bird_simpson = diversity(bird_mat, index = "simpson")
)

wild_div <- tibble(
  plot_id = rownames(wild_mat),
  wild_richness = specnumber(wild_mat),
  wild_shannon = diversity(wild_mat, index = "shannon"),
  wild_simpson = diversity(wild_mat, index = "simpson")
)

div_compare <- inner_join(bird_div, wild_div, by = "plot_id") %>%
  separate(plot_id, into = c("site", "plot"), sep = "_", remove = FALSE) %>%
  mutate(plot = as.numeric(plot))
```
#### 7b. diversity by days
```{r}
# Bird plot-day community
bird_day_comm <- birds_30min_day %>%
  count(site, plot, sample_day, species, name = "abundance") %>%
  unite("plot_day_id", site, plot, sample_day, remove = FALSE) %>%
  select(plot_day_id, species, abundance) %>%
  pivot_wider(names_from = species, values_from = abundance, values_fill = 0)

bird_day_mat <- bird_day_comm %>%
  column_to_rownames("plot_day_id")

bird_day_div <- tibble(
  plot_day_id = rownames(bird_day_mat),
  bird_richness = specnumber(bird_day_mat),
  bird_shannon = diversity(bird_day_mat, index = "shannon"),
  bird_simpson = diversity(bird_day_mat, index = "simpson")
)

# Wildlife plot-day community
wild_day_comm <- wildlife_independent_day %>%
  count(site, plot, sample_day, species, name = "abundance") %>%
  unite("plot_day_id", site, plot, sample_day, remove = FALSE) %>%
  select(plot_day_id, species, abundance) %>%
  pivot_wider(names_from = species, values_from = abundance, values_fill = 0)

wild_day_mat <- wild_day_comm %>%
  column_to_rownames("plot_day_id")

wild_day_div <- tibble(
  plot_day_id = rownames(wild_day_mat),
  wild_richness = specnumber(wild_day_mat),
  wild_shannon = diversity(wild_day_mat, index = "shannon"),
  wild_simpson = diversity(wild_day_mat, index = "simpson")
)

# Join paired plot-day diversity
div_compare_day <- inner_join(bird_day_div, wild_day_div, by = "plot_day_id") %>%
  separate(plot_day_id, into = c("site", "plot", "sample_day"), sep = "_", remove = FALSE) %>%
  mutate(
    plot = as.numeric(plot),
    sample_day = as.Date(sample_day)
  )
```
#### 8. Correlation tests
```{r}
# Richness
cor_rich_pearson <- cor.test(div_compare$bird_richness, div_compare$wild_richness, method = "pearson")
cor_rich_spearman <- cor.test(div_compare$bird_richness, div_compare$wild_richness, method = "spearman")

# Shannon
cor_shannon_pearson <- cor.test(div_compare$bird_shannon, div_compare$wild_shannon, method = "pearson")
cor_shannon_spearman <- cor.test(div_compare$bird_shannon, div_compare$wild_shannon, method = "spearman")

# Simpson
cor_simpson_pearson <- cor.test(div_compare$bird_simpson, div_compare$wild_simpson, method = "pearson")
cor_simpson_spearman <- cor.test(div_compare$bird_simpson, div_compare$wild_simpson, method = "spearman")
```
#### 9. Simple linear models and visualization
```{r}
lm_rich <- lm(bird_richness ~ wild_richness, data = div_compare)
lm_shannon <- lm(bird_shannon ~ wild_shannon, data = div_compare)
lm_simpson <- lm(bird_simpson ~ wild_simpson, data = div_compare)# Richness plot
ggplot(div_compare, aes(x = wild_richness, y = bird_richness)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~site, scales = "free") +
  labs(
    title = "Bird vs Wildlife Species Richness by Plot",
    x = "Wildlife richness",
    y = "Bird richness"
  ) +
  theme_bw()

# Shannon plot
ggplot(div_compare, aes(x = wild_shannon, y = bird_shannon)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~site, scales = "free") +
  labs(
    title = "Bird vs Wildlife Shannon Diversity by Plot",
    x = "Wildlife Shannon diversity",
    y = "Bird Shannon diversity"
  ) +
  theme_bw()

# Simpson plot
ggplot(div_compare, aes(x = wild_simpson, y = bird_simpson)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~site, scales = "free") +
  labs(
    title = "Bird vs Wildlife Simpson Diversity by Plot",
    x = "Wildlife Simpson diversity",
    y = "Bird Simpson diversity"
  ) +
  theme_bw()
```
#### 9b. Visualization by day
```{r}
# Bird vs Wildlife richness at plot-day level
ggplot(div_compare_day, aes(x = wild_richness, y = bird_richness)) +
  geom_point(alpha = 0.7, size = 1.8) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~site, scales = "free") +
  labs(
    title = "Bird vs Wildlife Species Richness by Shared Plot-Day",
    x = "Wildlife richness",
    y = "Bird richness"
  ) +
  theme_bw()

# Bird vs Wildlife Shannon at plot-day level
ggplot(div_compare_day, aes(x = wild_shannon, y = bird_shannon)) +
  geom_point(alpha = 0.7, size = 1.8) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~site, scales = "free") +
  labs(
    title = "Bird vs Wildlife Shannon Diversity by Shared Plot-Day",
    x = "Wildlife Shannon diversity",
    y = "Bird Shannon diversity"
  ) +
  theme_bw()

# Bird vs Wildlife Simpson at plot-day level
ggplot(div_compare_day, aes(x = wild_simpson, y = bird_simpson)) +
  geom_point(alpha = 0.7, size = 1.8) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~site, scales = "free") +
  labs(
    title = "Bird vs Wildlife Simpson Diversity by Shared Plot-Day",
    x = "Wildlife Simpson diversity",
    y = "Bird Simpson diversity"
  ) +
  theme_bw()
```